---
title: "Class Notes 2014-09-11"
author: 'R Pruim'
date: '2014-09-11'
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
```

## Old Faithful

### How long is a typical eruption? How much does the length of an eruption vary?

```{r}
require(MASS)   # geyser data lives here
mean(~duration, data=geyser)
median(~duration, data=geyser)

favstats( ~ duration, data=geyser )
```

### We need more than numbers

Looking at some plots shows us that the mean and median might not be very good summary numbers for this data.

```{r}
densityplot( ~ duration, data=geyser )
bwplot( ~ duration, data=geyser )
histogram( ~ duration, data = geyser, width = 0.2 )
```

There is also a quirky artifact in the data.

```{r}
histogram( ~ duration, data = geyser, width = 0.05 )
```

Apparently during the nigth, eruptions were listed as long or  short and recorded as 4 or 2 minutes.  See `?geyser` for more about the data.

### Duration vs wait

`waiting` is the length of time since the previous eruption.  For scatter plots, we put the explanatory variable (if there is one) on the x-axis and the response on the y-axis.  Since the eruptions come *after* the wait, we'll consider duration to be the response.

```{r}
xyplot( duration ~ waiting, data=geyser )
```

We can also calculate the time until the next eruption.  (You don't need to fully understand this code, but it is shifting the waits over one row.)

```{r}
geyser <- geyser %>% 
  mutate( time.til.next = c(tail(waiting, -1), NA) )
xyplot( time.til.next ~ duration, data=geyser )
```

This shows a much clearer picture.  After a longer eruption, we expect a longer wait.

### Z-scores

Z-scores measure how far above or below the mean a value is in units of standard deviations.  That is, they answer the question: 

> How many standard deviations above or below the mean is this value?  

Here's the formula:
$$
Z = \frac{\mathrm{value} - \mathrm{mean}}{\mathrm{standard\ deviation}}
$$

This gives us a standardized way of comparing values across distributions.

#### Example: Men's and Women's heights

Let's compare a 6'0" woman to a 6'5" man.  Which seems more unusual (i.e., farther above average)?

We can use this information to do the calculations.
```
 Height and Weight distributions

            height (male 18-74)     N(5'9",3") =    N(69,3)
            height (female 18-74)   N(5'3.5",2.5")= N(63.5,2.5)

            height (male 18-24)     N(5'10",3") =   N(70,2.8)
            height (female 18-24)   N(5'4.3",2.6")= N(64.3,2.6)

            height (male 11)        N(146cm,8cm) =  N(57.5,3.15)

            weight (male 18-24)     N(162,29.1)
            weight (female 18-24)   N(134,27)
```

```{r}
z_woman <- (72 - 64.3) / 2.6; z_woman
z_man <- (77 - 70) / 2.8; z_man
```
So it is more unusual for a woman to be 6'0" than for a man to be 6'5".

### Correlation coefficient

We would like a number that measures the strength (and direction) of linear associations between two variables.  The correleation coefficient ($r$) is computed by summing the products of z-scores and dividing by $n-1$.  We won't calculate this number by hand.

Here's what you need to know about $r$:

 * $r$ is always between -1 and 1
 * $r=1$ only for a perfect linear fit with positive slope
 * $r=-1$ only for a perfect linear fit with negative slope
 * we get the same number if we reverse the roles of explantory and response variables
 * the `cor()` function in R computes it for us


```{r}
# get the same value either way around for cor():
cor( waiting ~ duration, data=geyser )
cor( duration ~ waiting, data=geyser )
# this doesn't work because some we have some missing data
cor( time.til.next ~ duration, data=geyser )  
tail( geyser, 3 )
# ignore missing data -- use only complete observations
cor( time.til.next ~ duration, data=geyser, use="complete.obs") 
```

As expected, the correlation is stronger (closer to 1) for `time.til.next`.

#### Correlation Guessing Game

You can get calibrated to the correlation scale by playing this game:

 * [Correlation Guessing Game](http://www.istics.net/Correlations/)
 
