---
title: "More Randomization Distributions"
author: "R Pruim"
date: '2014-09-18'
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
```


## Survey Says

First, let's read in the data.  In RStudio,
there is a GUI for this in the environment tab.  (Look for Import Dataset.)  But we can't click on things in an RMarkdown document, so we need a command to do it.  `read.file()` can read from files or from URLs, so we can load the survey data directly from the course website.
```{r}
Survey <- read.file(
  "http://www.calvin.edu/~rpruim/courses/m145/F14/Data/Math145F14SurveyResults.csv", 
  stringsAsFactors=FALSE)
# turn things into factors if they don't have many different values
Survey <- factorize(Survey)  

# let's check that things look OK
summary(Survey)
head(Survey)
str(Survey)
```

Now we are ready to test the null hypothesis that the version
of the ticket question does not affect the answers.  That is, the proportion of subjects who say "yes" (they would buy a ticket) is the same whether they are told they lost a ticket or a $20 bill.

First, let's look at our data.
```{r}
tally( ticket ~ ticketVer, data=Survey)
# adding margins makes it clearer which proportions we are getting here
tally( ticket ~ ticketVer, data=Survey, margins=TRUE)
# If we prefer counts, we can do it this way
tally( ~ ticket + ticketVer, data=Survey)
tally( ticket ~ ticketVer, data=Survey, format="count")
```
Now let's compute our test statistic.  The difference in proportions is a good way to summarize our evidence.
```{r}
testStat <- diffprop( ticket ~ ticketVer, data=Survey)
testStat  # naming this makes it easier to use again later
```
The proportions are different in our two groups, but are they different enough to convince us that the version of the question makes a difference?  Or could this happen just by random chance?
Let's creat a randomization distribution to find out.

```{r}
Ticket.Rand <- do(10000) * 
  diffprop( ticket ~ shuffle(ticketVer), data=Survey)
head(Ticket.Rand)
histogram( ~ diffprop, data=Ticket.Rand, v=testStat )
# here is one way to do the two-sided version of the test
tally( ~( abs(diffprop) >= abs(testStat)), 
       data=Ticket.Rand, format="prop")
# here is a one-sided version
tally( ~( diffprop >= testStat), 
       data=Ticket.Rand, format="prop")
```
That's a pretty large p-value, so we can't reject the null hypothesis.
Our data are consistent with the hypothesis that the version of the question doesn't matter.  We say the difference we observed is **not statistically significant**.

This is not necessarily evidence in support of the null hypothesis -- we may simply not have enough data to tell one way or another.  We'll have to wait to find out more about evidence that can genuinely be said to support a null hypothesis.  Stay tuned...


