---
title: "Randomization Review"
author: "R Pruim"
date: '2014-09-19'
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
require(Lock5withR)
```


# Overview of Randomization Distributions we have seen

We have used randomization distributions in 5 situations so far.
The procedure for each is very similar:

1. State the Null and Alternative Hypotheses

  Remember these will be about a parameter or parameters.
  
2. Compute a Test Statistic

  Often the test statistic uses "data versions" of the parameters. If the hypothesis is about population means, the test statistics will involve sample means, etc.
  
3. Determine the p-value (from a simulated randomization distribution)

We need to simulate a world 

  a. where the null hypothesis is true,
    
  b. that reflects the design of our study, and 
    
  c. takes advantage of information the data provides about the population.
    
4. Interpret/Draw a conclusion

  Small p-values provide evidence against the null hypothesis.  
  When the p-value is *small enough* we say the results are **statistically significant**.  (The threshold for smallness depends on context.)

## Creating Randomization Distributions

Here is a side by side comparision of the ways we have created randomization distributions in our 5 key examples:

```{r eval=FALSE}
# rflip can simulate any proportion
Lady.Rand <- do(10000) * rflip(20)   # p = 0.50 by default
Hugo.Rand <- do(10000) * rflip(50, prob=1/6)
```
These three involve two variables and shuffle the explantory variable.
```{r eval=FALSE}
Taps.Rand <- do(10000) *
  diffmean( Taps ~ shuffle(Caffeine), data=CaffeineTaps)
Ticket.Rand <-do(10000) * 
  diffprop(ticket ~ shuffle(ticketVer), data=Survey)
Ticket.Rand <- do(10000) * 
  diffprop(ticket ~ shuffle(ticketVer), data=Survey)
```

Testing a mean is a little bit different because we want to 
shift the disribution to make the mean match the null hypothesis value.  This maintains the shape of the distribution.  Note the use of `resample()` here.  This is sampling *with replacement*.
```{r, eval=FALSE}
Temp.Rand <- do(10000) * 
  mean( ~ (BodyTemp + .34), data= resample(BodyTemp50)) 
```
Full details for each example appear below.

## Testing a Proprtion: 

### Lady Tasting Tea 

$H_0$: $p=0.50$; $n=20$

Test Stat: $x = 18$ or $\hat p = 18/20 =0.9$ 

```{r}
Lady.Rand <- do(10000) * rflip(20)
histogram ( ~ heads, data=Lady.Rand, v = 18)
# one-sided p-values
tally( ~ heads >= 18, data=Lady.Rand, format="prop")
tally( ~ prop >= 0.9, data=Lady.Rand, format="prop")
```

### Hugo

$H_0: p = 1/6$; $n=50$

Test Stat: $x = 16$ or $\hat p = 16/50 = 0.32$ 

```{r}
Hugo.Rand <- do(10000) * rflip(50, prob=1/6)
histogram ( ~ prop, data=Hugo.Rand, v = .32)
# one-sided p-values
tally( ~( heads >= 16), data=Hugo.Rand, format="prop")
tally( ~( prop >= 0.32), data=Hugo.Rand, format="prop")
```

### Testing the Difference Between Two Means:  Caffeine Taps

$H_0$: $\mu_1 = \mu_2$  ($\mu_1 - \mu_2 = 0$)

Test Stat:  $\overline x_1 - \overline x_2$  (Use `diffmean()`)

```{r}
testStat = diffmean(Taps ~ Caffeine, data=CaffeineTaps); testStat
Taps.Rand <- do(10000) *
  diffmean( Taps ~ shuffle(Caffeine), data=CaffeineTaps)
histogram( ~diffmean, data=Taps.Rand, v=testStat )
# two-sided
tally( ~(abs(diffmean) >= abs(testStat)), data=Taps.Rand, format="prop" )

```

### Testing the Difference Between Two Proportions: Survey Question

```{r, include=FALSE}
Survey <- read.file(
  "http://www.calvin.edu/~rpruim/courses/m145/F14/Data/Math145F14SurveyResults.csv", 
  stringsAsFactors=FALSE)
# turn things into factors if they don't have many different values
Survey <- factorize(Survey)  
```

$H_0$: $p_1 = p_2$ ($p_1 - p_2 = 0$)

Test Stat: $\hat p_1 = \hat p_2 = `r diffprop(ticket ~ ticketVer, data=Survey)`$

```{r}
testStat <- diffprop(ticket ~ ticketVer, data=Survey); testStat
Ticket.Rand <- do(10000) * 
  diffprop(ticket ~ shuffle(ticketVer), data=Survey)
histogram(~diffprop, data=Ticket.Rand, v=testStat)
# one-sided
tally(~ (diffprop >= testStat), data=Ticket.Rand, format="prop")
# one-side, alternative method
prop(~ (diffprop >= testStat), data=Ticket.Rand)
```

### Testing for Associaition Between Two Quantitative Variables: NFL Uniforms

$H_0$: $\rho = 0$

Test Stat: $r = `cor( ZPenYds ~ NFL_Malevolence, data=MalevolentUniformsNFL)`$

```{r}
head(MalevolentUniformsNFL)
xyplot( ZPenYds ~ NFL_Malevolence, data=MalevolentUniformsNFL)
testStat <- 
  cor( ZPenYds ~ NFL_Malevolence, data=MalevolentUniformsNFL)
testStat
NFL.Rand <- do(10000) * 
  cor(ZPenYds ~ shuffle(NFL_Malevolence), data=MalevolentUniformsNFL)
head(NFL.Rand)
histogram( ~ result, data=NFL.Rand)
# one-sided
tally( ~ (result >= testStat), data=NFL.Rand, format="prop")
# two-sided
tally( ~ (abs(result) >= testStat), data=NFL.Rand, format="prop")
```

### Testing a Mean: Body Temperature

$H_0$: $\mu = 98.6$

Test Stat:  $\overline x = `r mean(~BodyTemp, data=BodyTemp50)`$

```{r}
mean(~ BodyTemp, data=BodyTemp50)
histogram(~ BodyTemp, data=BodyTemp50)
98.6 - 98.26
# shift by 0.34 so the mean is like in the null hypothesis
Temp.Rand <- do(10000) * 
  mean(~ (BodyTemp + .34), data= resample(BodyTemp50)) 
histogram( ~ result, data=Temp.Rand, v=98.26)
tally( ~ result > 98.26, data=Temp.Rand, format="prop")
# a short cut
prop(~ result > 98.26, data=Temp.Rand)
```

