---
title: "Paired Designs"
author: "R Pruim"
date: '2014-09-23'
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
```

## Setting the Random Seed

Before we do our example, as a reminder, let's set the random seed so we get the "same random randomizations" every time we knit this file.

```{r}
set.seed(12345)   # we can choose any number here.
```

## Bowls of M and M's

A book by Tintle *et al* includes a data set based on an experiment to see whether univeristy students take more M and M's if they are presented with a larger bowl.
Each subject was "tested" twice (on different days), once with a large bowl and once with a small bowl.  Half of the
students had the large bowl the first time, the other half had the small bowl the first time.

```{r}
require(Tintle1)
head(BowlsMMs)
```
As we see, each case has two varialbes, the number of M & M's chosen form the large bowl and the number chosen from the smaller bowl.  Our null hypothesis is that on average, the difference between these two measurements is 0:

 * $H_0$: $\mu_{diff} = 0$.
 
 Our test statistic is the mean difference:
```{r}
mean( ~( Large - Small), data=BowlsMMs)
```

We can think about the randomization distribution at least two ways.  

### Converting this to a problem about one mean

Since our null hypothesis is about one mean (the mean difference), we could treat it just like the Body Temperature example.  But first we need to compute all those differences.

```{r}
MM2 <- mutate( BowlsMMs, diff = Large - Small )
head(MM2)
```
From this point on it is just like before
```{r}
testStat <- mean( ~diff, data=MM2 ); testStat

MM2.Rand <- do(10000) * 
  mean( diff - testStat, data=resample(MM2)) 
histogram( ~ result, MM2.Rand, v=testStat)
# 1-sided
tally( ~(result >= testStat), data=MM2.Rand, format="prop")
```
A 2-sided p-value would be about twice as large.

### Swapping the Small/Large labels

Another way to randomize -- a way that matches the design of 
the study well -- is to flip a coin to decide whether or not we randomly swap the value of `Small` with the value of `Large` for each subject.


Notice how some of the Small/Large values have been swapped 
in the example below.

```{r}
BowlsMMs
# let's randomly swap some
swap( BowlsMMs, Small ~ Large )
# here is a different randomization
swap( BowlsMMs, Small ~ Large )
```

Now we are all set to create a randomization distribution.
```{r}
BowlsMMs.Rand <- do(1000) *
  mean( Large - Small, data=swap(BowlsMMs, Small ~ Large))
histogram(~result, BowlsMMs.Rand, v=testStat)
tally( ~(result >= testStat), data=BowlsMMs.Rand, format="prop")
```
