---
title: "Birthdays"
author: 'R Pruim'
date: '2014-09-08'
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
require(lubridate)
```


## Looking at Birthdays Again

We have already looked at this plot a couple times

```{r}
xyplot( births ~ date, data=Births78) 
```

It was suggested that a possible explanation for the 
two parallel waves is a weekend/weekday affect.
We can check wether this is plausible by adding some features to our plot.

First let's color by day of week. The `wday` function in the 
lubridate pacakge will convert dates to day of the week for us.

```{r}
require(lubridate)
xyplot( births ~ date, data=Births78, 
        groups=wday(date)) 
```

This looks like some good evidence to support our conjecture.  There are primarily two colors in the lower
wave and 5 in the upper wave -- but there are some exceptions.  These are easier to spot if we "connect the dots".

```{r}
xyplot( births ~ date, data=Births78, 
        groups=wday(date), type='l' )
```

Now it is easier to see that there are a few days that
are "in the wrong group".  (These are the holidays.)

Finally, let's add a key so we know which day is which.
```{r}
xyplot( births ~ date, data=Births78, 
        groups=wday(date), type='l', 
        auto.key=TRUE)
```

Here's a much nicer version (fancier than you need to know how to 
do just now).

```{r}
xyplot( births ~ date, data=Births78, 
        groups=wday(date, label=TRUE, abbr=TRUE), type='l', 
        auto.key=list(columns=4, points=FALSE, lines=TRUE)
        )
```


We can also compute numbers and put them right into our text.  For example, the mean number of births is  `r mean(~births, data=Births78)`.   That's more digits than we need.   We could choose to round this to  `r round(mean(~births, data=Births78))`.  
