---
title: "Describing Distributions"
author: "R Pruim"
date: '2014-09-09'
output:
  pdf_document:
    fig_height: 3
    fig_width: 5
  html_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
```


## Pros and cons of Experimental vs Observational studies

 * Experiments may be artificial
 * Observation studies need to be wary about confounding
 * Some experiments are not ethical
 


## Distributions

When we talk about the distribution of a variable, we are  talking about the answers to the following questions: 

  * what values? 
  * how often?

We can summarize distributions in several ways:

 * tables 
 * graphs 
 * numerical summaries
 * adjectives that describe the "shape"

### HELPrct

The Health Evaluation and Linage to Primary Care data set has information collected on individuals admitted for treatment of an adiction to alcohol, cocaine, or heroin.

#### substance

```{r}
tally( ~ substance, data=HELPrct)
tally( ~ substance, data=HELPrct, format="prop")
bargraph( ~ substance, data=HELPrct)
```

#### sex
```{r}
tally( ~ sex, data=HELPrct, format="percent")
```

#### age
```{r}
favstats( ~age, data=HELPrct )
```


#### i1

`i1` is the average number of drinks (standard units) consumed per day over the 30 days before admission to the program.

```{r}
favstats( ~ i1, data=HELPrct )
bwplot( ~i1, data=HELPrct)
histogram( ~ i1, data=HELPrct )
```

```{r}
bwplot( substance ~ i1, data=HELPrct )
histogram( ~ i1 | substance, data=HELPrct, width=20 )
favstats( ~ i1 | substance, data=HELPrct )
favstats(i1 ~ substance, data=HELPrct )
densityplot( ~ i1, groups=substance, data=HELPrct, width=20, auto.key=TRUE, lwd=3 )
```


