---
title: "Math 145 Week 3 Q & A"
author: "R Pruim"
date: "Fall 2014"
output:
  html_document:
    fig_height: 3
    fig_width: 5
  pdf_document:
    fig_height: 3
    fig_width: 5
  word_document:
    fig_height: 3
    fig_width: 5
---


```{r include=FALSE}
# Don't delete this chunk if you are using the mosaic package
# This loads the mosaic and dplyr packages
require(mosaic)
```

```{r include=FALSE}
# Some customization.  You can alter or delete as desired (if you know what you are doing).

# This changes the default colors in lattice plots.
trellis.par.set(theme=theme.mosaic())  

# knitr settings to control how R chunks work.
require(knitr)
opts_chunk$set(
  tidy=FALSE,     # display code as typed
  size="small"    # slightly smaller font for code
)
# This loads the mosaic data sets.  (Could be deleted if you are not using them.)
require(mosaicData)   
```


**Student:**
Hello Professor,

I am having difficulty getting a tally function to work. I am trying to make my own randomization distribution based on criteria from question 4.118 on page 277 in the text book. When I try to tally for heads more than .5 it tells me that there are 1 true and 0 false which isn't right. here is my rcode.

```{r}
Rand.Rand <- do(5000)*rflip(100, prob=.5)
histogram(~heads, data=Rand.Rand, width=1)
mean(~heads, data=Rand.Rand)
tally(~heads>.5, data=Rand.Rand, format="prop")
```

If you could help that would be great and thank you.

**Prof Pruim:**
As is often the case, R is doing exactly what you told it to do.  You just didn't ask the right question.  You are asking R to tally up which random runs have more than 0.5 heads.  They **all** do.  

Basically, you are mixing counts and proportion. You either want `~(prob > 0.5)` or `(~heads > 50)`.  For example,

```{r}
tally( ~(prop > 0.5), data=Rand.Rand, format="prop" )
```
Alternatively, you could do
```{r}
prop( ~(prop > 0.5), data=Rand.Rand)
```
